Parsing Postfix Log Files

### 2009-04-26 doc/logparser-implementation.tex
Changes based on proofreading chapters 4 and 5.

This chapter documents the implementation of \parsername{}, a parsed based
on the architecture described in \sectionref{parser architecture}.  Any
design may look good when examined in the abstract, but the real test of
the design comes with the first concrete implementation; only then do any
practical difficulties come to light.  Implementing this architecture was
relatively straightforward: the difficulties came not from the architecture
and design, but from anomalies in the log files and Postfix's behaviour, as
described in \sectionref{complications}.  \parsername{} successfully deals
with all the difficulties that were discovered during its development, but
further difficulties may arise during future usage; unfortunately solving
all potential difficulties that may arise is an intractable task, but the
descriptions of the solutions developed thus far should help if someone
needs to develop a solution of their own.

\parsername{} deals with all the eccentricities and oddities of parsing
Postfix log files, presenting the resulting data in a normalised, simple to
use representation.  Unfortunately, dealing with the complications that
arise sometimes requires the parser to discard log lines (e.g.\
\sectionref{aborted delivery attempts}), and, less frequently, to discard a
data structure (e.g.\ \sectionref{timeouts during data phase}).
\parsername{} can parse log files from Postfix version 2.2 through to
version 2.5, and should be able to parse log files from later versions with
only minor modifications or an updated ruleset.  The parser's task is to
follow the journey each mail takes through Postfix, combining the data
captured by rules into a coherent whole, saving it in a useful and
consistent form, and performing housekeeping duties.  The intermingling of
log lines from different mails immediately rules out the possibility of
handling each mail in isolation; the parser must handle multiple mails in
parallel, each potentially at a different stage in its journey, without any
interference between mails --- except in the minority of cases where
intra-mail interference is required, e.g.\ mail re-injected for forwarding
(\sectionref{Re-injected mails}).  The best way to deal with intermingling
of log lines is to maintain state information for every unfinished mail,
and to manipulate the appropriate mail correctly for each log line
encountered.

@@

The next three sections document the implementation of the three components
of the architecture.  First is the framework, including its initialisation
phase, the parsing process, and the conveniences it offers to users of
\parsername{}; that section finishes with the features that are important
for this thesis but not strictly necessary for parsing: the performance
data collected by the framework, the optimisations that can be disabled to
show their effect, and the debugging options it provides.  The second
component of the architecture is the actions, starting with a graph of how
often each action is specified by rules, a description of why some actions
are more popular than others, and how this popularity has no influence on
how often an action is invoked.  Every action that is part of \parsername{}
is documented, and the actions section concludes with a description of the
process of adding a new action.  Rules are the final component of the
architecture, and also the most visible to advanced users, e.g.\ systems
administrators, because it is likely that they will need to add their own
rules to recognise their log lines.  An example rule used by \parsername{}
is examined, with every field clearly documented; that is followed by a
description of adding new rules, and how to determine the value of each of
the fields.  \parsername{} provides a utility to create regexes from
unrecognised log lines; the algorithm it uses is documented, including the
differences between it and the original algorithm it is based on.  The
rules section finishes with a discussion of how \parsername{} uses rule
conditions and overlapping rules, plus a description of the regex snippets
provided to ease the process of writing regexes.

@@

    \item The parser acts as a second source of information about Postfix's
        operation, based on empirical evidence from log files.  A separate
        project could compare the empirical knowledge inherent in
        \parsername{} with Postfix's documentation and source code to see
        how closely the two agree.

@@

The flow chart in \figureref{flow chart image} shows the most common paths
a connection or mail can take through \parsername{}; decision boxes and the
difficulties described in \sectionref{complications} are excluded for the
sake of clarity.  The flow chart is intended to be a graphical overview of
how a mail progresses through both Postfix and \parsername{}, providing an
overall context into which the detailed descriptions in the remainder of
this chapter will fit, in particular the actions (\sectionref{actions in
implementation}) and complications (\sectionref{complications}).  The
states and state transitions shown in the flow chart will be explained
later in this section.

@@

    \item The client disconnects \flowchart{DISCONNECT, TIMEOUT}{6}, either
        normally or with an error.  If Postfix has rejected any
        \acronym{SMTP} commands the data gathered from those rejections
        will be saved to the database; if there were no rejections there
        will not be any data to save.  Any mails accepted will already have
        been cloned so their data is in another data structure, and will be
        delivered in the same way as mails that entered the system by any
        other route.

@@

    \item The mail will have one or more delivery attempts
        \flowchart{MAIL\_SENT, SAVE\_DATA}{8}.

@@

    \item The mail is removed from the Postfix queue.  This is the last log
        line for this particular mail, though it may be indirectly referred
        to if it was re-injected.  The mail is cleaned up and entered in
        the database, then deleted from the state tables
        \flowchart{COMMIT}{11}.

@@

A database is used to store both the rules and the data gleaned by parsing
Postfix log files.  Understanding the database schema is important in
understanding the actions of the parser, and essential to developing
further applications that utilise the data; \sectionref{database as API}
describes how the database schema functions as an \acronym{API}.

The database schema can be conceptually divided in two: the rules used to
recognise log lines, and the data saved from the parsing of log files.
Each rule has a regex to recognise log lines and capture data from them,
and specifies the action to be invoked when a log line is recognised; they
also have several fields that aid the user in understanding the meaning of
the log lines recognised by each rule.  The rules are described in detail
in \sectionref{rules in implementation}, but the rules table is documented
in \sectionref{rules table} with the rest of the database schema.

The data saved from parsing the log files is divided into two tables:
connections and results.  The connections table contains an entry for every
mail accepted and every connection that rejected a delivery attempt; the
individual fields will be described in \sectionref{connections table}.  The
results table will have at least one entry for each entry in the
connections table; its fields will be covered in detail in
\sectionref{results table}.  A diagram of the database schema is provided
in \figureref{Diagram of the database schema picture}, to complement the
in-depth descriptions of each table.

An important but easily overlooked benefit of storing the rules in the
database is the link between rules and results: if more information is
required when examining a result, the rule that produced the result is
available for inspection because each result references the rule that
created it.  No ambiguity is possible about which rule resulted in a
particular result, eliminating one potential source of confusion.

A clear, comprehensible schema is essential when using the data extracted
from log lines; it is more important when using the data than when storing
it, because storing the data is a once-off operation, whereas utilising
the data requires frequent searching, sorting, and manipulation of the data
to produce customised reports or statistics.

@@

            \item New columns can be added to the tables used by the
                program, using DEFAULT clauses or TRIGGERS to populate
                them.

@@

Rules are discussed in detail in \sectionref{rules in implementation}, but
the structure of the rules table is documented here alongside the other
tables in the database.  Rules are created by the user, and will not be
modified by \parsername{}, except when it updates the hits and hits\_total
fields.  Rules recognise the individual log lines, capturing data to be
saved in the connections and results tables, and specifying the action to
invoke for each recognised log line.

@@

    \item [id] A unique identifier that other tables can use when referring
        to a specific rule.

@@

    \item [hits\_total] The total number of log lines recognised by this
        rule over all runs of the parser; hits is reset to zero each time
        the parser is run, but hits\_total is not.  This field will be set
        by the parser rather than the rule author.

@@

Every accepted mail and every connection that rejected a mail delivery
attempt will have a single entry in the connections table containing all of
the following below.  For an incoming connection, the client is the remote
machine, and the server is the local machine; for outbound mail delivery
attempts, the roles are reversed.

@@

    \item [id] This field uniquely identifies the row.

@@

    \item [helo] The hostname used in the HELO command.  The HELO hostname
        occasionally changes during a connection, presumably because spam
        or virus senders think it is a good idea.  By default Postfix only
        logs the HELO hostname when it rejects a mail delivery attempt, but
        it is quite simple to rectify this as described in
        \sectionref{logging helo}.

@@

Every recognised log line will have a row in the results table, associated
with a single connection; a single connection will have at least one result
associated with it, but will usually have several, and may have hundreds.

@@

    \item [id] A unique identifier for this result.

@@

    \item [smtp\_code] The \acronym{SMTP} code associated with the log
        line.  In general an \acronym{SMTP} code is only present in a log
        line representing a rejection or final delivery; results whose log
        line did not contain an \acronym{SMTP} code will duplicate the
        \acronym{SMTP} code of other results in that connection.  Some
        final delivery log lines do not contain an \acronym{SMTP} code,
        e.g.\ when Postfix delivers to a user's mailbox; in those cases the
        \acronym{SMTP} code is specified by the rule's
        \texttt{result\_data} field, based on the success or failure
        represented by the log line.

@@

    \item [sender] The sender's email address.  This may change during a
        connection if the client uses different sender addresses for
        multiple rejected delivery attempts; however there will not be
        different sender addresses in the results for one accepted mail.

@@

    \item [message\_id] The message-id of the accepted mail, again present
        for delivered mails only.

@@

The role of the framework in this architecture was described in detail in
\sectionref{framework in architecture}; this section is concerned with the
implementation of the framework within \parsername{}.  The framework
manages the parsing process, taking care of the drudgery and boring tasks,
provides services to the actions, and implements some efficiency measures.

Each time the parser is run the framework performs some initialisation
tasks, setting up several state tables that will later be used by actions.
Data about the connections and mails being processed is held in
\texttt{connections} and \texttt{queueids} respectively.  The other state
tables are used when working around complications in parsing:
\texttt{timeout\_queueids} is used when dealing with connections that time
out during the DATA phase (\sectionref{timeouts during data phase});
\texttt{bounce\_queueids} is part of the solution to bounce notification
mails being delivered before their creation is logged (\sectionref{Bounce
notification mails delivered before their creation is logged}); and
\texttt{postsuper\_deleted\_queueids} caches information about mails that
were recently deleted by the administrator so that subsequent log lines
processed by the \action{SAVE\_DATA} action can be discarded
(\sectionref{Mail deleted before delivery is attempted}).

@@

The framework is now ready to begin parsing.  For each log line it will use
those rules whose \texttt{program} field equals the program in the log line
(previously described in \sectionref{rule conditions in implementation}),
falling back to generic rules if necessary, and finally warning if the log
line is unrecognised.  The repetitive nature of log files gives them high
compression ratios; the framework uses a Perl module named
\texttt{IO::Uncompress::AnyUncompress} to read compressed log files, saving
users the trouble of uncompressing them to a temporary file before parsing
begins.  When used interactively, the framework displays a progress bar to
show how far parsing has progressed through the log file and how long the
remainder of the parsing process is likely to take.  The progress bar is
not as accurate when parsing compressed log files, because a compressed
block uncompresses to a variable number of log lines; variation between the
recognition and processing time of individual log lines also affects its
accuracy, but overall the progress bar is a useful addition.  \parsername{}
is primarily intended for parsing complete log files, but with minor
modifications it could be used to parse a live log file, periodically
checking it for new log lines; this could be very useful for programs that
work best with up to date data, e.g.\ a program for monitoring the health
of the mail system, or graphs showing activity over the last five minutes.

@@

    \item The number of log lines and rules used for each Postfix component
        (\graphref{Mean number of rules used per log line for each Postfix
        component}).

@@

    \item The rule ordering used can be changed from optimal (the default,
        most efficient) to shuffled (intended to represent an unordered
        ruleset) or reverse (reverse of optimal, least efficient).  The
        results of this are shown in \sectionref{rule ordering for
        efficiency}.

@@

    \item Each regex is compiled and the result cached when the ruleset is
        loaded; this optimisation can be disabled and the regex compiled
        every time it is used when recognising log lines.  The effect of
        this optimisation is shown in \sectionref{Caching compiled
        regexes}.

@@

The framework also provides several debugging options, to aid in writing or
correcting rules, or figuring out why the parser is not behaving as
expected.  In increasing order of severity they are:

@@

    \item Each result can be extended with extra debugging information,
        which is useful when a warning dumps a mail's data for inspection.
        The extra information added is: the log line's timestamp in human
        readable form; the entire log line; the name of the log file and
        the number of the log line within it.  This information is not
        stored in the database.

@@

A user, typically a mail administrator, would use these options when having
difficulty extending the ruleset to recognise log lines not recognised by
\parsernames{} \numberOFrules{} rules.  Option 1 is useful for debugging a
single rule, to check if it is recognising log lines it should not;
figuring out that a rule is not recognising log lines that it should
recognise is a more difficult task.  Option 2 is useful when \parsername{}
is warning about parsing problems, because extra information about a mail
will be present in the warning message.  Option 3 is less useful because of
the volume of data it produces, though for small log file snippets, e.g.\
1000 log lines, it is possible to manually verify that the correct rule
recognised each log line.  Option 4 is not useful for most users: it is for
debugging problems with the data that is saved to the database, and
sufficient safeguards are in place that there should be plenty of warning
messages explaining what is wrong without using this option.

@@

Actions are the component of this architecture responsible for processing
all of the inputs recognised by rules; in \parsername{} they reconstruct
the journey each mail takes through Postfix, dealing with all the
complications and difficulties that arise.  \parsername{} has
\numberOFactions{} actions and \numberOFrules{} rules, with an uneven
distribution of rules to actions as shown in \graphref{Distribution of
rules per action}.  Unsurprisingly, the action with the most associated
rules is \action{DELIVERY\_REJECTED}, the action that processes Postfix
rejecting a mail delivery attempt; next is \action{SAVE\_DATA}, the action
that saves useful information without doing any other processing.  The
third most common action is, perhaps surprisingly, \action{UNINTERESTING}:
this action does nothing when invoked, allowing uninteresting log lines to
be parsed without any effects; it does not imply that the input is
ungrammatical or unparsed.  Generally rules specifying the
\action{UNINTERESTING} action recognise log lines that are not associated
with a specific mail, e.g.\ notices about configuration files changing;
these log lines are recognised and processed so that the framework can warn
about unrecognised log lines, informing the user that they need to extend
the ruleset.  Most of the remaining actions have only one or two associated
rules, because they will only ever have one or two log line variants, e.g.\
all log lines showing that a remote client has connected are recognised by
a single rule and processed by the \action{CONNECT} action.

No correlation exists between how often an action is specified by rules and
how often an action is invoked because a rule recognises a log line.
\Graphref{Number of times each action was invoked excluding 22 and 62--68}
shows the number of times each action was invoked when parsing the
\numberOFlogFILES{} log files used to generate the statistics in
\sectionref{parser efficiency}, \textit{excluding\/} log files 22 \&
62--68, because their content is extremely skewed by two mail loops.  The
action most commonly specified by rules, \action{DELIVERY\_REJECTED}, is
the third most commonly invoked action; the most commonly invoked actions,
\action{CONNECT} and \action{DISCONNECT}, are each specified by only one
rule.  The \action{UNINTERESTING} action, the third most commonly specified
action, is halfway down the graph.

As expected, some actions have been invoked almost exactly the same number
of times: almost every \action{CONNECT} will have a \action{DISCONNECT},
with only a 0.00042\% difference between the number of times the two were
invoked; every mail that enters Postfix's queue will be managed by
\daemon{qmgr} (\action{MAIL\_QUEUED}) and processed by \daemon{cleanup}
(\action{CLEANUP\_PROCESSING}), and again the two have only a 0.01648\%
difference between their number of invocations.
\action{CLEANUP\_PROCESSING} is invoked slightly more often than
\action{MAIL\_QUEUED}: occasionally, an accepted mail in the process of
being transferred is interrupted by a timeout, and there is a log line from
\daemon{cleanup} but not from \daemon{qmgr}, as described in
\sectionref{discarding cleanup log lines}.

@@

This section documents the actions found in \parsername{}; it may help to
revisit the flow chart in \sectionref{flow chart} to see how a mail passes
from one action to another as its log lines are recognised.  The words
\textit{mail\/} and \textit{connection\/} are used in the actions
descriptions below because they are less unwieldy and more specific than
\textit{state table entry\/}; a connection becomes a mail during the
\action{CLONE} action, which processes Postfix accepting a delivery
attempt, and the data structure moves from the connections state table to
the queueids state table.

@@

If the log line has enough information to identify the correct connection
or mail, each action will save all the data captured by the recognising
rule's regex; usually, log lines that lack identifying information will be
processed by the \action{UNINTERESTING} action.  Each action is passed the
same arguments:

@@

@@

    \item [BOUNCE\_CREATED] Postfix 2.3 and subsequent versions log the
        creation of bounce messages, and this action processes those log
        lines.  This action creates a new mail if necessary; if the mail
        already exists the unknown origin flag will be removed from it.
        The mail will be marked as a bounce notification.  To deal with
        complication \sectionref{Bounce notification mails delivered before
        their creation is logged}, this action checks a cache of recent
        bounce mails, to avoid incorrectly creating bounce mails when log
        lines are out of order.

@@

    \item [COMMIT] Enter the data from the mail into the database.  Entry
        will be postponed if the mail is a child waiting to be tracked
        (\sectionref{Re-injected mails}).  Once entered in the database,
        the mail will be usually be deleted from the state tables, but
        deletion will be postponed if the mail is the parent of mail
        re-injected for forwarding (\sectionref{Re-injected mails}).

@@

    \item [SAVE\_DATA] Every action that can locate the correct entry in
        the state tables saves any data captured by the recognising rule's
        regex to it.  The \action{SAVE\_DATA} saves data in this way but
        does not do anything else; it is invoked for log lines that contain
        useful data but do not require any further processing.

@@

Adding new actions is not as facile as adding new rules, though care has
been taken in the architecture and implementation to make adding new
actions as painless as possible; one of the few limitations is that
\parsername{} is written in Perl, so new actions must also be written in
Perl.  The implementer writes a subroutine that accepts the standard
arguments given to actions, and registers it with the framework by calling
the framework's \texttt{add\_actions()} subroutine.  The new action must be
registered before the rules are loaded, because it is an error for a rule
to specify an unregistered action; this helps catch mistakes made when
adding new rules.  No other work is required from the implementer to
integrate the action into the parser; all of their attention and effort can
be focused on correctly implementing their new action.  The action may need
to extend the list of valid combinations described in \sectionref{out of
order log lines} if the new action creates a different set of acceptable
programs, but this would only be necessary if the new action processes log
lines from Postfix components that \parsername{} does not have rules for,
e.g.\ \daemon{virtual} or \daemon{lmtp}.

@@

    \item [result\_data and connection\_data:] are used to provide data not
        present in the log line, and are unused in this rule.

>>>>

This chapter documents the implementation of \parsername{}, a parser for
Postfix log files based on the architecture described in \sectionref{parser
architecture}.  Any design may look good when examined in the abstract, but
the real test of the design comes with the first concrete implementation;
only then do any practical difficulties come to light.  Implementing this
architecture was straightforward --- the difficulties came not from the
architecture, but from anomalies in the log files and Postfix's behaviour,
as described in \sectionref{complications}.  \parsername{} successfully
deals with all the difficulties that were discovered during its
development, but further difficulties may arise during future usage;
unfortunately solving all potential difficulties that may arise is an
impossible task, but the descriptions of the solutions developed thus far
should help if someone needs to solve a problem of their own.

\parsername{} deals with all the eccentricities and oddities of parsing
Postfix log files, presenting the resulting data in a normalised, simple to
use representation.  Unfortunately, dealing with the complications that
arise sometimes requires the parser to discard log lines (e.g.\
\sectionref{aborted delivery attempts}), and, less frequently, to discard a
data structure (e.g.\ \sectionref{timeouts during data phase}).
\parsername{} can parse log files from Postfix version 2.2 through to
version 2.5, and should be able to parse log files from later versions with
only minor modifications or an updated ruleset.  The parser's task is to
follow the journey each mail takes through Postfix, combining the data
captured by rules into a coherent whole, and saving it in a useful and
consistent form.  The intermingling of
log lines from different mails immediately rules out the possibility of
handling each mail in isolation; the parser must handle multiple mails in
parallel, each potentially at a different stage in its journey, without any
interference between mails --- except in the minority of cases where
intra-mail interference is required, e.g.\ mail re-injected for forwarding
(\sectionref{Re-injected mails}).  The best way to deal with intermingling
of log lines is to maintain state information for every unfinished mail,
and to manipulate the appropriate mail correctly for each log line
encountered.

@@

The next three sections document the implementation of the three components
of the architecture.  First is the framework, including its initialisation
phase, the parsing process, and the conveniences it offers to users of
\parsername{}; that section finishes by describing features that are
important for this thesis but not strictly necessary for parsing: the
performance data collected by the framework, the optimisations that can be
disabled to show their effect, and the debugging options the framework
provides.  The second component of the architecture is the actions,
starting with a graph showing how often each action is specified by rules,
a description of why some actions are more popular than others, and how
this popularity has no influence on how often an action is invoked.  Every
action that is part of \parsername{} is documented, and the actions section
concludes with a description of the process of adding a new action.  Rules
are the third component of the architecture, and also the most visible to
advanced users, e.g.\ systems administrators, because it is likely that
they will need to add their own rules to recognise their own log lines.  A
sample rule used by \parsername{} is examined, with every field clearly
documented; that is followed by a description of adding new rules, and how
to determine the value of each of the new rule's fields.  \parsername{}
provides a utility to create regexes from unrecognised log lines; the
algorithm it uses is documented, including the differences between it and
the original algorithm it is based on.  The rules section finishes with a
discussion of how \parsername{} uses rule conditions and overlapping rules,
plus a description of the regex snippets provided to ease the process of
writing regexes.

@@

    \item The parser acts as a second source of information about Postfix's
        operation, based on empirical evidence gathered from log files.  A
        separate project could compare the empirical knowledge inherent in
        \parsername{} with Postfix's documentation and source code to see
        how closely the two agree.

@@

The flow chart in \figureref{flow chart image} shows the most common paths
a connection or mail can take through \parsername{}; decision boxes and the
difficulties described in \sectionref{complications} are excluded for the
sake of clarity.  The flow chart is intended to be a graphical overview of
how a mail progresses through both Postfix and \parsername{}, providing an
overall context into which the detailed descriptions in the remainder of
this chapter will fit, in particular the actions (\sectionref{actions in
implementation}) and complications (\sectionref{complications}).  The
states and state transitions shown in the flow chart will be explained
in this section.

@@

    \item The client disconnects \flowchart{DISCONNECT \textnormal{or}
        TIMEOUT}{6}, either normally or with an error.  If Postfix has
        rejected any mail delivery attempts the data gathered from those
        rejections will be saved to the database \flowchart{DISCONNECT}{7};
        if there were no rejections there will not be any data to save.
        Any mails accepted will already have been cloned so their data is
        in another data structure, and will be delivered in the same way as
        mails that entered the system by any other route.

@@

    \item The mail will have one or more delivery attempts
        \flowchart{MAIL\_SENT \textnormal{or} SAVE\_DATA}{8}.

@@

    \item The mail is removed from the Postfix queue.  This is the last log
        line for this particular mail, though it may be indirectly referred
        to if it was re-injected.  The mail is cleaned up and entered in
        the database, then deleted from the state tables
        \flowchart{COMMIT}{10}.

@@

An \acronym{SQL} database is used to store both the rules and the data
gleaned by parsing Postfix log files.  Understanding the database schema is
helpful in understanding the actions of the parser, and essential to
developing further applications that utilise the data; \sectionref{database
as API} describes how the database schema functions as an \acronym{API}.

The database schema can be conceptually divided in two: the rules used to
recognise log lines, and the data saved from the parsing of log files.
Each rule has a regex to recognise log lines and capture data from them,
and specifies the action to be invoked when a log line is recognised; they
also have several other fields used by the parser, and several fields that
aid the user in understanding the meaning of the log lines recognised by
each rule.  The rules are described in detail in \sectionref{rules in
implementation}, but the rules table is documented in \sectionref{rules
table} with the rest of the database schema.

The data saved from parsing log files is divided into two tables:
connections and results.  The connections table contains an entry for every
mail accepted and every connection that rejected a delivery attempt; the
individual fields will be described in \sectionref{connections table}.  The
results table will have at least one entry for each entry in the
connections table; its fields will be covered in detail in
\sectionref{results table}.  A diagram of the database schema is provided
in \figureref{Diagram of the database schema picture}, to complement the
in-depth descriptions of each table.

An important but easily overlooked benefit of storing the rules in the
database is the link between rules and results: if more information is
required when examining a result, the rule that produced the result is
available for inspection because each result references the rule that
created it.  No ambiguity is possible about which rule or action created a
particular result, eliminating one potential source of confusion.

A clear, comprehensible schema is essential when using the data extracted
from log lines; it is more important when using the data than when storing
it, because storing the data is a once-off operation, whereas utilising
the data requires frequent searching, sorting, and manipulation of the
data.

@@

            \item New columns can be added to tables, using DEFAULT clauses
                or TRIGGERS to populate them.

@@

Rules are discussed in detail in \sectionref{rules in implementation}, but
the structure of the rules table is documented here alongside the other
tables in the database.  Rules are created by the user, and will not be
modified by \parsername{}, except when it updates the hits and hits\_total
fields.  Rules recognise individual log lines, capturing data to be saved
in the connections and results tables, and specifying the action to invoke
for each recognised log line.

@@

    \item [id] A unique identifier that other tables can use when referring
        to a specific rule.  This will be assigned by the database.

@@

    \item [hits\_total] The total number of log lines recognised by this
        rule over all runs of the parser; hits starts from zero each time
        the parser is run, but hits\_total is not.  This field will be set
        by the parser rather than the rule author.

@@

Every accepted mail and every connection that rejected a mail delivery
attempt will have a single entry in the connections table containing all of
the fields below.  For an incoming connection, the client is the remote
machine, and the server is the local machine; for outbound mail delivery
attempts, the roles are reversed.

@@

    \item [id] This field uniquely identifies the row.  This will be
        assigned by the database.

@@

    \item [helo] The hostname used in the HELO command.  The HELO hostname
        occasionally changes during a connection, presumably because spam
        or virus senders think it is a good idea.  Postfix only logs the
        HELO hostname when it rejects a mail delivery attempt, but it is
        quite simple to rectify this as described in \sectionref{logging
        helo}.

@@

Every recognised log line will have a row in the results table, and each
row is associated with a single connection; a single connection will have
at least one result associated with it, but will usually have several, and
may have hundreds.

@@

    \item [id] A unique identifier for this result.  This will be assigned
        by the database.

@@

    \item [smtp\_code] The \acronym{SMTP} code associated with the log
        line.  In general an \acronym{SMTP} code is only present in a log
        line representing a mail being delivered or a mail delivery attempt
        being rejected; results whose log line did not contain an
        \acronym{SMTP} code will duplicate the \acronym{SMTP} code of other
        results in that connection.  Some mail delivery log lines do not
        contain an \acronym{SMTP} code, e.g.\ when Postfix delivers to a
        user's mailbox; in those cases the \acronym{SMTP} code is specified
        by the rule's \texttt{result\_data} field, based on the success or
        failure represented by the log line.

@@

    \item [sender] The sender's email address.  This may change during a
        connection if the client uses different sender addresses for
        multiple rejected delivery attempts; however, the results for one
        accepted mail will only have one sender address.

@@

    \item [message\_id] The message-id of the accepted mail, again it is
        present for delivered mails only.

@@

The role of the framework in this architecture was described in detail in
\sectionref{framework in architecture}; this section is concerned with the
implementation of the framework within \parsername{}.  The framework
manages the parsing process, taking care of the drudgery and boring tasks,
provides services to the actions, and implements some optimisations.

The framework performs some initialisation tasks each time the parser is
run, setting up several state tables that will later be used by actions.
Data about the connections and mails being processed is held in
\texttt{connections} and \texttt{queueids} respectively.  The other state
tables are used when solving complications that arise during parsing:
\texttt{timeout\_queueids} is used when dealing with connections that time
out during the DATA phase (\sectionref{timeouts during data phase});
\texttt{bounce\_queueids} is part of the solution to bounce notification
mails being delivered before their creation is logged (\sectionref{Bounce
notification mails delivered before their creation is logged}); and
\texttt{postsuper\_deleted\_queueids} caches information about mails that
were recently deleted by the administrator, so that subsequent log lines
processed by the \action{SAVE\_DATA} action can be discarded
(\sectionref{Mail deleted before delivery is attempted}).

@@

The framework is now ready to begin parsing.  For each log line it will use
those rules whose \texttt{program} field equals the program in the log line
(described in \sectionref{rule conditions in implementation}), falling back
to generic rules if necessary, and finally warning if the log line is
unrecognised.  The repetitive nature of log files gives them high
compression ratios; the framework uses a Perl module named
\texttt{IO::Uncompress::AnyUncompress} to read compressed log files, saving
users the trouble of uncompressing them to a temporary file before parsing
begins.  When used interactively, the framework displays a progress bar to
show how far parsing has progressed through the log file and how long the
remainder of the parsing process is likely to take.  The progress bar is
not as accurate when parsing compressed log files, because each compressed
block uncompresses to a variable number of log lines; variation between the
recognition and processing time of individual log lines also affects its
accuracy, but overall the progress bar is a useful addition.  \parsername{}
is primarily intended for parsing complete log files, but with minor
modifications it could be used to parse a live log file, periodically
checking it for new log lines; this could be very useful for programs that
work best with up to date data, e.g.\ a program for monitoring the health
of the mail system, or graphs showing activity over the last five minutes.

@@

    \item The number of log lines and rules used per log line for each
        Postfix component (\graphref{Mean number of rules used per log line
        for each Postfix component}).

@@

    \item The rule ordering used can be changed from optimal (the default,
        most efficient) to shuffled (intended to represent an unordered
        ruleset) or reverse (reverse of optimal, least efficient).  The
        results of parsing with the three different orderings are shown in
        \sectionref{rule ordering for efficiency}.

@@

    \item Each regex is compiled and the result cached when the ruleset is
        loaded; this optimisation can be disabled and the regex compiled
        every time it is used when recognising log lines.  The effect this
        optimisation has is addressed shown in \sectionref{Caching compiled
        regexes}.

@@

The framework also provides several debugging options, to aid in writing or
correcting rules, or figuring out why \parsername{} is not behaving as
expected.  In increasing order of severity they are:

@@

    \item Each result can be extended with extra debugging information,
        which is useful when a warning dumps a data structure for
        inspection.  The extra information added is: the log line's
        timestamp in human readable form; the entire log line; the name of
        the log file and the line number of the log line within it.  This
        extra information is not stored in the database.

@@

A user, typically a mail administrator, would use these options when having
difficulty extending the ruleset to recognise log lines not recognised by
\parsernames{} \numberOFrules{} rules.  Option 1 is useful for debugging a
single rule, to check if it is recognising log lines it should not;
figuring out that a rule is not recognising log lines that it should
recognise is a more difficult task.  Option 2 is useful when \parsername{}
is warning about parsing problems, because extra information about a mail
will be present in the warning message.  Option 3 is less useful because of
the volume of data it produces, though for small log file snippets, e.g.\
1000 log lines, it is possible to manually verify that the correct rule
recognised each log line.  Option 4 is not useful for most users: it is for
debugging problems within \parsername{}, and sufficient safeguards are in
place that there should be plenty of warning messages explaining what is
wrong without using this option.

@@

Actions are the component of this architecture responsible for processing
all of the inputs recognised by rules; in \parsername{} they reconstruct
the journey each mail takes through Postfix, dealing with all the
complications and difficulties that arise.  \parsername{} has
\numberOFactions{} actions and \numberOFrules{} rules, with an uneven
distribution of rules to actions as shown in \graphref{Distribution of
rules per action}.  Unsurprisingly, the action with the most associated
rules is \action{DELIVERY\_REJECTED}, the action that processes Postfix
rejecting a mail delivery attempt; next is \action{SAVE\_DATA}, the action
that saves useful information without doing any other processing.  The
third most common action is, perhaps surprisingly, \action{UNINTERESTING}:
this action does nothing when invoked, allowing uninteresting log lines to
be parsed without any effects; it does not imply that the input is
ungrammatical or unparsed.  Generally, rules specifying the
\action{UNINTERESTING} action recognise log lines that are not associated
with a specific mail, e.g.\ notices about configuration files changing;
these log lines are recognised and processed so that the framework can warn
about unrecognised log lines, informing the user that they need to augment
the ruleset.  Most of the remaining actions have only one or two associated
rules, because that input category will only ever have one or two log line
variants, e.g.\ all log lines showing that a remote client has connected
are recognised by a single rule and processed by the \action{CONNECT}
action.

No correlation exists between how often an action is specified by rules and
how often an action is invoked because a rule recognises a log line.
\Graphref{Number of times each action was invoked excluding 22 and 62--68}
shows the number of times each action was invoked when parsing the
\numberOFlogFILES{} log files used to generate the statistics in
\sectionref{parser efficiency}, \textit{excluding\/} log files 22 \&
62--68, because their contents are extremely skewed by two mail loops,
described in \sectionref{Characteristics of the 93 log files}.  The action
most commonly specified by rules, \action{DELIVERY\_REJECTED}, is the third
most commonly invoked action; the most commonly invoked actions,
\action{CONNECT} and \action{DISCONNECT}, are each specified by only one
rule.  The \action{SAVE\_DATA} and \action{UNINTERESTING} actions, the
second and third most commonly specified actions respectively, are halfway
down the graph.

As should be expected, some actions have been invoked almost exactly the
same number of times: almost every \action{CONNECT} will have a
\action{DISCONNECT}, with only a 0.00042\% difference between the number of
times the two were invoked; every mail that enters Postfix's queue will be
managed by \daemon{qmgr} (\action{MAIL\_QUEUED}) and processed by
\daemon{cleanup} (\action{CLEANUP\_PROCESSING}), and again the two have
only a 0.01648\% difference between their number of invocations.
\action{CLEANUP\_PROCESSING} is invoked slightly more often than
\action{MAIL\_QUEUED}: occasionally, an accepted mail in the process of
being transferred is interrupted by a timeout, and there is a log line from
\daemon{cleanup} but not from \daemon{qmgr}, as described in
\sectionref{discarding cleanup log lines}.

@@

This section documents the actions found in \parsername{}; it may help to
revisit the flow chart in \sectionref{flow chart} to see how a mail passes
from one action to another as its log lines are recognised.  The words
\textit{mail\/} and \textit{connection\/} are used in the action
descriptions below because they are less unwieldy and more specific than
\textit{state table entry\/}; a connection becomes a mail during the
\action{CLONE} action, which processes Postfix accepting a delivery
attempt, and the data structure is copied from the connections state table
to the queueids state table.

@@

If the log line has enough information to identify the correct connection
or mail, each action will save all the data captured by the recognising
rule's regex to the connection or mail; usually, log lines that lack
identifying information will be processed by the \action{UNINTERESTING}
action.  Each action is passed the same arguments:

@@

\noindent{}\textbf{\parsernames{} actions:}

@@

    \item [BOUNCE\_CREATED] Postfix 2.3 and subsequent versions log the
        creation of bounce messages, and this action processes those log
        lines.  This action creates a new mail if necessary; if the mail
        already exists the unknown origin flag will be removed from it.  To
        deal with complication \sectionref{Bounce notification mails
        delivered before their creation is logged}, this action checks a
        cache of recent bounce mails, to avoid incorrectly creating bounce
        mails when log lines are out of order, and also marks the mail as a
        bounce notification.

@@

    \item [COMMIT] Enter the data from the mail into the database.  Entry
        will be postponed if the mail is a child waiting to be tracked
        (\sectionref{Re-injected mails}).  Once entered into the database,
        the mail will be usually be deleted from the state tables, but
        deletion will be postponed if the mail is the parent of mail
        re-injected for forwarding (\sectionref{Re-injected mails}).

@@

    \item [SAVE\_DATA] Every action that can locate the correct data
        structure in the state tables saves any data captured by the
        recognising rule's regex to it.  The \action{SAVE\_DATA} saves data
        in this way but does not do anything else; it is invoked for log
        lines that contain useful data but do not require any further
        processing.

@@

Adding new actions is not as simple as adding new rules, though care has
been taken in the architecture and implementation to make adding new
actions as painless as possible; one of the few limitations is that
\parsername{} is written in Perl, so new actions must also be written in
Perl.  The implementer writes a subroutine that accepts the standard
arguments given to actions, and registers it with the framework by calling
the framework's \texttt{add\_actions()} subroutine.  No other work is
required from the implementer to integrate the action into the parser; all
of their attention and effort can be focused on correctly implementing
their new action.  The action may need to extend the list of valid
combinations described in \sectionref{out of order log lines} if the new
action creates a different set of acceptable programs, but this would only
be necessary if the new action processes log lines from Postfix components
that \parsername{} does not have rules for, e.g.\ \daemon{virtual} or
\daemon{lmtp}.  The new action must be registered before the rules are
loaded, because it is an error for a rule to specify an unregistered
action; this helps catch mistakes made when adding new rules.

@@

    \item [result\_data and connection\_data:] are used to provide data not
        present in the log line, but are unused in this rule.
